Open Data, Open Source and Open Standards in chemistry: \\ The Blue Obelisk five years on

### 2011-05-31 paper.tex
More updates from PMR

[Few lines on Quixote here]

@@

@@

\subsection*{Chemical Databases}

@@

@@

(Say something here about benchmarks to measure accuracy. Clear examples
 of performance on open datasets are required. Otherwise there is
 nothing to counter anecdotal evidence or FUD spread by others.)

(Better engagement with industry...make it clear what and how industry
 members can engage with projects.)

@@

>>>>

The Quixote project epitomises the full use of the Blue Obelisk
software and is described in detail in a sibling
article. Here we observe that it is possible to convert legacy files
of all sorts into semantic chemistry and extract
those parts which are suitable for input to computational chemistry
programs. This chemistry is then combined with
generic concepts of computational chemistry ({\it e.g.} strategy,
machine resources, timing, accuracy etc.) into the
legacy inputs for a wide range of programs. Quixote itself follows
Blue Obelisk principles in that it does not manage
the submission and monitoring of jobs but resumes action when the jobs
have been completed, and then applies a range
of parsing and transformation tools to create standardised semantic
chemical content. A major feature of Quixote is
that it requires all concepts to validate against dictionaries and the
process of parsing files necessarily generates
communally-agreed dictionaries, which represent an important step
forward in the Open specifications for Blue Obelisk.
When widely-deployed, Quixote will advertise the value of Open
community standards for semantics to the world.

The Quixote project is not dependent on any particular technology,
other than the representation of computational
chemistry in CML and the management of semantics through CML
dictionaries. At present, we use JUMBO-Converters for most
 of the semantic conversion, Lensfield2 for the workflow and Chempound
(chem\#) to store and disseminate the results.

@@

In May 2011, the EBI ran an industry-oriented workshop (MIOSS -
Molecular Informatics Open Source Software). This explored the role of
Open Source in industrial laboratories and companies, and several of
the presenters are among the authors of this paper.
The meeting identified that Open Source was extremely valuable to
industry, not just because it is 'free as in beer'
but because it allows the validation of source code, data and
computational procedures. A phrase from the meeting
summed it up: "The ice is beginning to melt", signifiying that we can
expect a rapid increase in industry's interest
in Open Source.

Some of the discussion was on business models. There are difficulties
for software for which there is no formal
transaction, other than downloasding and agreeing to license terms.
COmpanies are concerened about training and support
and in some cases, product liability. One anecdote was of a company
which wished to donate money to an Open Source
project but could not find a mechanism to do so.

There is a considerable amount of contribution-in-kind, both from
enhancements to software and also completely new
systems and toolkits, and companies are finding it easier to create
mechanisms for releasing Open Source software
without violating confidentiality or incurring liability.

@@

\subsection*{Chemical database software}

@@

Besides the traditional and proven relational databse approach, with
added chemical features ('cartridges'), there is growing
interest in tools and approaches based on the web philosophy and
practice. Several groups are experimenting with RDF
on the assumption that generic high-performance solutions are
appearing. RDF allows everything to be described by
URIs (data, molecules, dictionaries, relations). The Chempound system,
as deployed in
Quixote and elsewhere, is an RDF-based approach to chemical structures
and compounds and their properties. For small
to medium-sized collections (such as an individual's calculations or
literature retrieval), there are many RDF tools
(e.g. SIMILE, Apache Jena) which can operate in machine memory and provide
the flexibility that RDF offers. For larger
systems, it is unclear whtether complete RDF solutions (e.g. Virtuoso)
will be satisfactory or whether a hybrid system
based on name-value pairs (e.g. CouchDB, MongoDB) will be sufficient.

@@

@@

\subsection*{The Blue Obelisk Data Repository, BODR;
http://sourceforge.net/projects/bodr/}

From the beginning, the Blue Obelisk created a repository of key data,
particularly that which would be used in
algorithms and where there was a need to ensure that the values were
standard between codes. Examples are atomic
masses and conversion between physical constants. In principle, this
material can be copied between sites such as IUPAC and
wikipedia and continued practice in this area should lead to a
enhancement in the quality of community reference data.

\subsection*{NMRShiftDB}

NMRShiftDB represents one of the earliest (2002?) resources for Open
community-contributed data. Groups which
measure NMR spectra or extract it from the literature contribute to
NMRShiftDB which provides an Open resource where
entries can be searched by chemical structure or properties
(especially peaks). It is, not surprisingly, difficult to
extract large amounts of altruistic contribution (as happens in
wikipedia) but it is increasingly possible to link
data capture with data publication. For example, the Blue Obelisk has
enough software that it is possible to create
a seamless chain for converting NMR structures in-house into
NMRShiftDB entries. If and when the chemistry community
encourages or requires semantic publication of spectra, rather than
PDFs, it would be possible to populate NMRShiftDB rapidaly
along the the lines of CrystalEye.

\subsection*{CrystalEye}

CrystalEye represents the cost-effective extraction of data from the
literature where this is published both
Openly and semantically. The cost is extremely low and software can be
run every night, updating the entries
(currently at about 250,000 structures). CrystalEye serves as a model
for a high-value, high-quality Open data
resource, including the licensing of each component as
Panton-compatible Open data.
